Bayesian Inference for Species Distribution Modelling with Gaussian Processes

### 2021-04-19 _book/manuscript.tex
more more more

\emph{Metropolis-Hastings MCMC} is one of the most used algorithms for sampling from the posterior distribution to perform a Bayesian update.
While it is not always the fastest in practice and requires manual tuning to work effectively, Metropolis-Hastings provides a foundation for more complex algorithms such as \emph{Hamiltonian Monte Carlo} \citep{Brooks2011} with \emph{No-U-Turn sampling} \citep{Homan2014}.
To reiterate, the purpose of sampling in Bayesian inference is to draw from some density for parameters \(\theta\).
In this case, the density we are drawing from is the Bayesian posterior \(P(\theta|data)\), which for the purposes of MCMC is referred to as the \emph{target distribution}.
The basic idea of Metropolis-Hastings, to return to the previously mentioned example of approximating \(I=\int h(x)f(x)dx\), is to construct a Markov chain \(X_0,...,\) with a stationary distribution of \(h\).
Assuming the sequence has sufficiently many observations, the law of large numbers applies and the following convergence holds.
\[\frac{1}{N}\sum_{i=0}^{N}w(X_{i})\xrightarrow{P} E(w(X))=I\]
The Metropolis-Hastings algorithm provides a method for sampling from the \emph{stationary distribution} of a Markov chain designed in such a way that this distribution approximates the posterior distribution.
Initially, the probability of realizing a state as part of the sampled sequence is skewed, but importantly, over time the sequence will converge.
Critically, the law of large numbers guarantees that with sufficiently many iterations of Metropolis-Hastings, the generated sequence of states will converge on this stationary distribution.
Essentially, given some sequence of states \(X_{0},X_{1},...,X_{i}\) from a Markov-Chain, where \(X_{0}\) is chosen arbitrarily, an iteration of Metropolis-Hastings produces the next state to include in the sequence.
As presented formally in Definition \ref{def:methas}, this is achieved by generating a \emph{proposal} for \(X_{i+1}\) from the \emph{proposal distribution} \(q(y|X_{i})\), transition kernel, and then accepting the proposal with some probability dependent on the relative, target probabilities of the current state \(X_{i}\) and the proposal \citep{Wasserman2004}.
\begin{definition}[Metropolis-Hastings MCMC]
\protect\hypertarget{def:methas}{}{\label{def:methas} \iffalse (Metropolis-Hastings MCMC) \fi{} }\emph{\(X_{i+1}\) is generated given \(X_{0},X_{1},...,X_{i}\) in the following manner:}

@@

While Metropolis-Hastings may seem appealing because it is both memoryless and able to approximate very complex distributions, its downfall lies in the manual tuning necessary to achieve convergence of the stationary distribution to the target distribution in a reasonable number of iterations.
Recall that Metropolis-Hastings generates proposals by sampling from some distribution.
In the special case of random walk Metropolis-Hastings, this is a normal distribution centered about \(X_{i}\).
Here, the standard deviation of the normal distribution, known as the step size, dictates the relative distance in the sample space between \(X_{i}\) and the generated proposal.
If the step size is too low, the algorithm will make very small steps and may miss key features of the target distribution, causing the stationary distribution to require a much larger number of iterations to converge on the target distribution.
On the other hand if the step size is too large, proposals will be overwhelmingly generated from the low-probability tails of the distribution, again resulting in a lack of convergence and poorly representative samples from the stationary distribution.
While some methods have been proposed for automatically tuning the step size parameter \citep{Graves2011}, iterating on complex, Bayesian models to tune a parameter is not particularly efficient nor computationally feasible in many cases.
For this reason, other approximation algorithms are used in practical implementations of Bayesian inference.
One such algorithm, \emph{Hamiltonian Monte Carlo} (HMC), relies on theorhetical physics to compute a latent momentum variable which is applied to a hamiltonian, effectively simulating a ball rolling around the multi-dimensional sample space \citep{Brooks2011}.
Despite the increased computational cost, this method is appealing because of its ability to generate proposals from distant regions of the stationary distribution with high acceptance probabilities.
This means that in practice, the HMC algorithm's stationary distribution converges on the target distribution with far fewer iterations than traditional Metropolis-Hastings implementations.
Though HMC still requires the user to specify a step size as well as a number of steps to move the hamiltonian before considering a proposal, in practice it is much more efficient and requires fewer iterations on a model.
Additionally, a proposed extension to HMC called the \emph{No-U-Turn Sampler} (NUTS) automatically determines the number of steps and was empirically demonstrated to perform at least equally as well as standard HMC \citep{Homan2014}.
The approximation algorithms breifly covered in this chapter underpin modern applications for Bayesian inference, including Stan, which are discussed in the following chapter.

\hypertarget{modern-bayesian-inference-with-stan}{%
\chapter{Modern Bayesian inference with Stan}\label{modern-bayesian-inference-with-stan}}

Co-founded at Columbia University by Bob Carpenter and Andrew Gelman, Stan is a domain specific programming language that provides tools for specifying, executing and evaluating statistical models.
Currently comprised of more than 45 core developers from a wide array of educational institutions, private entities, and research organizations, over the past 9 years Stan has grown to become one of the leading options for performing Bayesian inference computationally \citep{Stan2021}.
Among the many capabilities of the Stan language, is the ability to define priors, specify dependencies between variables, and impose known limits on parameters.
After all model components have been specified, a Stan model compiles to C++ and can then be executed to produce a fit.

@@

While Stan is not strictly a tool for Bayesian statistical methods, it is most often used for its efficient MCMC sampler for Bayesian inference.
In essence, Stan has a sampler that takes a model and creates a separate program which explores the sample space to learn the parameters of interest in the modelling task using what is essentially the previously mentioned NUTS HMC sampling algorithm to generate draws from the poseterior distribution.
As it is written in C++ and thus produces compiled code, a Stan program is able to take advantage of the full capabilities of the machine on which it is executed.
Theoretically, Stan allows computer scientists and statisticians to construct, evaluate, and iterate on extremely complex, hierachical models in one consolidated environment.

A typical Stan program is comprised of several, optional \emph{program blocks}, which provide a general organizational structure to a Stan program and generally allow for the declaration of variables and accompanying statements about those variables.
Block types include functions, data, transformed data, parameters, transformed prarameters, model, and generated quantities.
Each of these block types serve a different purpose in a general modelling schema; however no block is required for the Stan compiler to execute, and an empty string is considered a valid Stan program (though it may not produce a very informative model).
After the program samples the user specified number of iterations and completes execution, a fit object and accompanying summary statistics are produced, which allow for performing post-hoc evaluation such as the posterior predictive check.

>>>>

\emph{Metropolis-Hastings MCMC} is a standard algorithm for sampling from the posterior distribution to perform a Bayesian update.
While it is not always the fastest in practice and requires manual tuning to work effectively, Metropolis-Hastings provides a foundation for more complex algorithms such as \emph{Hamiltonian Monte Carlo} \citep{Brooks2011} with \emph{No-U-Turn sampling} \citep{Homan2014}.
To reiterate, the purpose of sampling in Bayesian inference is to draw from the posterior density with a goal of estimating \(P(\theta|data)\) for parameters \(\theta\).
The Bayesian posterior \(P(\theta|data)\), for the purposes of MCMC, is also referred to as the \emph{target distribution}.
The basic idea of Metropolis-Hastings, to return to the previously mentioned example of approximating \(I=\int h(x)f(x)dx\), is to construct a Markov chain \(X_0,...,\) with a stationary distribution of \(h\).
Assuming the sequence has sufficiently many observations, the law of large numbers applies and the following convergence holds.
\[\frac{1}{N}\sum_{i=0}^{N}w(X_{i})\xrightarrow{P} E(w(X))=I\]
The Metropolis-Hastings algorithm provides a method for sampling from the \emph{stationary distribution} of a Markov chain designed in such a way that this distribution approximates the posterior distribution.
Initially, the probability of realizing a state as part of the sampled sequence is skewed, but importantly, over time the sequence will converge.
Critically, the law of large numbers guarantees that with sufficiently many iterations of Metropolis-Hastings, the generated sequence of states will converge on this stationary distribution.
Essentially, given some sequence of states \(X_{0},X_{1},...,X_{i}\) from a Markov-Chain, where \(X_{0}\) is chosen arbitrarily, an iteration of Metropolis-Hastings produces the next state to include in the sequence.
As presented formally in Definition \ref{def:methas}, this is achieved by generating a \emph{proposal} for \(X_{i+1}\) from the \emph{proposal distribution} \(q(y|X_{i})\) and then accepting the proposal with some probability dependent on the relative, target probabilities of the current state \(X_{i}\) and the proposal \citep{Wasserman2004}.
\begin{definition}[Metropolis-Hastings MCMC]
\protect\hypertarget{def:methas}{}{\label{def:methas} \iffalse (Metropolis-Hastings MCMC) \fi{} }\emph{\(X_{i+1}\) is generated given \(X_{0},X_{1},...,X_{i}\) in the following manner:}

@@

While Metropolis-Hastings may seem appealing because it is both memoryless and able to approximate very complex distributions, its downfall lies in the manual tuning of the \emph{step size} parameter necessary to achieve convergence of the stationary distribution to the target distribution in a reasonable number of iterations.
Recall that Metropolis-Hastings generates proposals by sampling from some distribution.
In the special case of \emph{random walk Metropolis-Hastings}, this is a normal distribution centered about \(X_{i}\).
Here, the standard deviation of the normal distribution, known as the step size, dictates the relative distance in the sample space between \(X_{i}\) and the generated proposal.
If the step size is too low, the algorithm will make very small steps and may miss key features of the target distribution, causing the stationary distribution to require a much larger number of iterations to converge on the target distribution.
On the other hand if the step size is too large, proposals will be overwhelmingly generated from the low-probability tails of the distribution, again resulting in a lack of convergence and poorly representative samples from the stationary distribution.
While some methods have been proposed for automatically tuning the step size parameter \citep{Graves2011}, iterating on complex, Bayesian models to tune a parameter is not particularly efficient nor computationally feasible in many cases.
For this reason, other approximation algorithms are used in practical implementations of Bayesian inference.
One such algorithm, \emph{Hamiltonian Monte Carlo} (HMC), relies on theorhetical physics to compute a latent momentum variable which is applied to a hamiltonian, effectively simulating a ball rolling around the multi-dimensional sample space \citep{Brooks2011}.
Despite the increased computational cost, this method is appealing because of its ability to generate proposals from distant regions of the stationary distribution with high acceptance probabilities.
This means that in practice, the HMC algorithm's stationary distribution converges on the target distribution with far fewer iterations than traditional Metropolis-Hastings implementations.
Though HMC still requires the user to specify a step size as well as a number of steps to move the hamiltonian before considering a proposal, in practice it is much more efficient and requires fewer iterations on a model.
Additionally, a proposed extension to HMC called the \emph{No-U-Turn Sampler} (NUTS) automatically determines the number of steps and was empirically demonstrated to perform at least equally as well as standard HMC \citep{Homan2014}.
The approximation algorithms breifly covered in this chapter underpin modern applications for Bayesian inference, including Stan, which are discussed in the following chapter.

\hypertarget{bayesian-inference-with-stan}{%
\chapter{Bayesian inference with Stan}\label{bayesian-inference-with-stan}}

Created at Columbia University by Bob Carpenter and Andrew Gelman, Stan is a domain specific programming language that provides tools for specifying, executing and evaluating statistical models.
Currently comprised of more than 45 core developers from a wide array of educational institutions, private entities, and research organizations, over the past 9 years the Stan project has grown to become one of the leading options for performing Bayesian inference computationally \citep{Stan2021}.
Among the many capabilities of the Stan language, is the ability to define priors, specify dependencies between variables, and impose known limits on parameters.
After all model components have been specified, a Stan model compiles first to C++ and then to binary, which can then be executed to produce an estimation of the posterior densities of all parameters, including intermediate ones, obtained through sampling, with accompanying metadata for evaluating the sampling process.

@@

While Stan is not strictly a tool for Bayesian statistical methods, it is most often used for its efficient MCMC sampler for Bayesian inference.
In essence, Stan has a sampler that takes a model and creates a separate program which explores the sample space to learn the parameters of interest in the modelling task using what is essentially the previously mentioned NUTS HMC sampling algorithm to generate draws from the poseterior distribution.
As it is written in C++ and thus produces binary code, a Stan program is able to take advantage of the full capabilities of the machine on which it is executed.
Theoretically, Stan allows computer scientists and statisticians to construct, evaluate, and iterate on extremely complex, hierachical models in one consolidated environment.

A typical Stan program is comprised of several, optional \emph{program blocks}, which provide a general organizational structure to a Stan program and generally allow for the declaration of variables and accompanying statements about those variables.
Block types include functions, data, transformed data, parameters, transformed prarameters, model, and generated quantities.
Each of these block types serve a different purpose in a general modelling schema; however no block is required for the Stan compiler to execute, and an empty string is considered a valid Stan program (though it may not yield a very useful model).
After the program samples the user specified number of iterations and completes execution, a fit object and accompanying summary statistics are produced, which allow for performing post-hoc evaluation such as the posterior predictive check.

### 2021-04-19 index.Rmd
more more more

--- 
title: "Bayesian Inference for Species Distribution Modelling with Gaussian Processes"
subtitle: "Isaac William Caruso"
author: |
  | Amherst College
  | Department of Computer Science
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
description: "Bayesian Inference for Species Distribution Modelling with Gaussian Processes"
documentclass: book
link-citations: yes
bibliography:
  - references.bib
  - packages.bib
site: bookdown::bookdown_site
biblio-style: apalike
fontsize: 12pt
linestretch: 1.7
geometry: "left=1in, right=1in, top=1in, bottom=1in"
---
```{r setup, include = FALSE}
library(tidyverse)
library(mosaic)
library(mdsr)
library(knitr)
library(kableExtra)
library(latex2exp)
library(extrafont)
library(rstan)
library(mdsr)
#font_import(pattern = "cmu*") 
loadfonts()
setwd("~/bayes-gp-sdm-manuscript/")
```
# Probability and Bayes’ theorem

@@

```{example montecarl}
Suppose we have two binomially distributed r.vs---$X \sim Binom(n,p_1)$, $Y \sim Binom(m,p_2)$---and want to evaluate $\delta = p_2 - p_1$. 
This problem can be analyzed using Bayesian inference on a beta-binomial model, but would require evaluating complex integrals over the joint posterior of $X$ and $Y$ to obtain the density.
Monte Carlo Integration allows for estimating an integral using repeated sampling from the two independent posterior densities---$P_1$ and $P_2$--- and plotting samples using a histogram provides a qualitative estimate for the parameter of interest, $\delta$ [@Wasserman2004].
In simplest terms the beta prior distribution $Beta(\alpha, \beta)$ (the proper conjugate prior for the binomial likelihood) could be a flat prior $\alpha = \beta = 1$, on the parameters $p_1$ and $p_2$ [@Bloom2014].
$f(p_1,p_2|X,Y)=f(p_1|X)f(p_2|Y)$, because the two posteriors $f(p_1|X)$ and $f(p_2|Y)$ are conditionally independent given the data.
Thus the resultant posteriors are beta distributions, as follows [@Bloom2014].
$$D_1 = Beta(X+1,n-X+1)$$
$$D_2 = Beta(Y+1,m-Y+1)$$
Monte Carlo integration simulates the quantity $\delta$ by drawing $(P_1^{(1)},P_2^{(2)}),...,(P_1^{(N)},P_2^{(N)})$ from the independent beta posteriors $D_1$ and $D_2$ for $i=1,...,N$ assuming N is sufficiently large, every $\delta^{(i)}=P_2^{(i)} - P_1^{(i)}$.
Accordingly, $\delta$ is approximated using the previously established equivalency, $$\delta \approx \frac{1}{N} \sum_i^N \delta^{(i)} = \frac{1}{N} \sum_i^N (P_2^{(i)} - P_1^{(i)})$$ [@Wasserman2004].
To provide a concrete example, if the number of trials is 20, $n=m=20$, $X = 11$, and $Y = 4$ Sampling 10000 times from the in the previously described manner yields the histogram presented in Figure \@ref(fig:mon) 
```
```{r mon, fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histogram displaying displaying posterior of delta from Monte Carlo integration"}
p1 <- rbeta(10000, 12, 10)
p2 <- rbeta(10000, 5, 17)
delta <- p2 - p1
delta_df <- data.frame(delta = delta)
hist <- ggplot(delta_df, aes(x=delta)) + 
  geom_histogram(colour="black", fill="grey") + #+#, fill="grey") +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank(),
    text = element_text(size=15, family="CMU Serif"))
hist
```

@@

Created at Columbia University by Bob Carpenter and Andrew Gelman, Stan is a domain specific programming language that provides tools for specifying, executing and evaluating statistical models.
Currently comprised of more than 45 core developers from a wide array of educational institutions, private entities, and research organizations, over the past 9 years Stan has grown to become one of the leading options for performing Bayesian inference computationally [@Stan2021].
Among the many capabilities of the Stan language, is the ability to define priors, specify dependencies between variables, and impose known limits on parameters.
After all model components have been specified, a Stan model compiles to C++ and can then be executed to produce a fit.

@@

A typical Stan program is comprised of several, optional *program blocks*, which provide a general organizational structure to a Stan program and generally allow for the declaration of variables and accompanying statements about those variables.
Block types include functions, data, transformed data, parameters, transformed prarameters, model, and generated quantities. 
Each of these block types serve a different purpose in a general modelling schema; however no block is required for the Stan compiler to execute, and an empty string is considered a valid Stan program (though it may not produce a very useful model).
After the program samples the user specified number of iterations and completes execution, a fit object and accompanying summary statistics are produced, which allow for performing post-hoc evaluation such as the posterior predictive check. 
```{r include=FALSE, cache=TRUE}
sharks = read.csv('./data/florida-shark.csv')
```
```{css, echo=FALSE}
.small {
  font-size: small;
}
```

@@

sharks_data = list(
    'N' = length(sharks$population),
    'population'= sharks$population,
    'year'= sharks$year,
    'attacks' = sharks$attacks
)
fit = stan(model_code = sharks_stancode, cores=4, data=sharks_data, chains=4, iter=2000)
```
```{r, cache=TRUE, class.output="small", comment='', tidy=TRUE, size='footnotesize', tidy.opts = list(blank = FALSE, width.cutoff = 30), echo=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)
print(summary(fit, pars=c('alpha', 'beta'))$summary, digits=2)
```
Bob notes that beta, the rate parameter he is interested in understanding, seems to be increasing by nearly 3% each year, although the 97.5% confidence interval of 3.5% causes Bob to wonder how certain he should be in his conclusion about beta. To get a better understanding of the relative frequencies of different samples of the model's parameter values, Bob plots the sample values for each iteration on a histogram, as presented in Figure \@ref(fig:hist).
```{r hist, cache=TRUE, comment='', fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histograms displaying frequencies of varying sampled values for parameters alpha and beta"}
stan_hist(fit, pars=c('alpha', 'beta')) +#, fill="grey") +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank()) +
  labs(xaxis = c('alpha', 'beta'))
```
The final analysis step for Bob's Bayesian model is ensuring the chains mixed properly, which Bob determines qualitatively using a trace plot as shown in Figure \@ref(fig:trace). 
```{r trace, cache=TRUE, comment='', fig.height=2, echo=FALSE, warning=FALSE, fig.align='center', fig.cap="Each chain represents a sequence of states sampled from the posterior for parameters alpha and beta. Four chains were sampled for this model, and this trace plot is a qualitative method for determining if the sample space for each parameter was sufficiently explored. Here, the sampled value of each parameter (y-axis) is plotted for every iteration Stan evaluated. In this case, Bob specified 2000 total iterations and 1000 burn in iterations, so the first 1000 samples were discarded and not included in the posterior or the trace."}
stan_trace(fit, pars=c('alpha', 'beta')) +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank())
```
As it is clear the chains mixed well, Bob is now relatively confident in the results his model is producing. While he does trust the samples evaluated by his model and his analysis of Figure \@ref(fig:hist)---that the rate of shark attacks is indeed increasing in Florida by some amount---the general lack of concentrated density in both histograms leads Bob to determine that he should collect more data and continue sampling his fit in an attempt to gain a better understanding of preciesly how much shark attacks are increasing each year.
Stan, and more generally Bayesian inference, is a good option for a task like this, because Bob can simply collect more data and continue training this model with his new data, without the need to start the fitting process over again.

>>>>

--- 
title: "Bayesian Inference for Species Distribution Modelling with Gaussian Processes"
subtitle: "Isaac William Caruso"
author: |
  | Amherst College
  | Department of Computer Science
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
description: "Bayesian Inference for Species Distribution Modelling with Gaussian Processes"
documentclass: book
link-citations: yes
bibliography:
  - references.bib
  - packages.bib
site: bookdown::bookdown_site
biblio-style: apalike
fontsize: 12pt
linestretch: 1.7
geometry: "left=1in, right=1in, top=1in, bottom=1in"
---
```{r setup, include = FALSE}
library(tidyverse)
library(mosaic)
library(mdsr)
library(knitr)
library(kableExtra)
library(latex2exp)
library(extrafont)
library(rstan)
library(mdsr)
font_import(pattern = "Latin Modern Roman*") 
loadfonts()
setwd("~/bayes-gp-sdm-manuscript/")
```
# Probability and Bayes’ theorem

@@

```{example montecarl}
Suppose we have two binomially distributed r.vs---$X \sim Binom(n,p_1)$, $Y \sim Binom(m,p_2)$---and want to evaluate $\delta = p_2 - p_1$. 
This problem can be analyzed using Bayesian inference on a beta-binomial model, but would require evaluating complex integrals over the joint posterior of $X$ and $Y$ to obtain the density.
Monte Carlo Integration allows for estimating an integral using repeated sampling from the two independent posterior densities---$P_1$ and $P_2$--- and plotting samples using a histogram provides a qualitative estimate for the parameter of interest, $\delta$ [@Wasserman2004].
In simplest terms the beta prior distribution $Beta(\alpha, \beta)$ (the proper conjugate prior for the binomial likelihood) could be a flat prior $\alpha = \beta = 1$, on the parameters $p_1$ and $p_2$ [@Bloom2014].
$f(p_1,p_2|X,Y)=f(p_1|X)f(p_2|Y)$, because the two posteriors $f(p_1|X)$ and $f(p_2|Y)$ are conditionally independent given the data.
Thus the resultant posteriors are beta distributions, as follows [@Bloom2014].
$$D_1 = Beta(X+1,n-X+1)$$
$$D_2 = Beta(Y+1,m-Y+1)$$
Monte Carlo integration simulates the quantity $\delta$ by drawing $(P_1^{(1)},P_2^{(2)}),...,(P_1^{(N)},P_2^{(N)})$ from the independent beta posteriors $D_1$ and $D_2$ for $i=1,...,N$ assuming N is sufficiently large, every $\delta^{(i)}=P_2^{(i)} - P_1^{(i)}$.
Accordingly, $\delta$ is approximated using the previously established equivalency, $$\delta \approx \frac{1}{N} \sum_i^N \delta^{(i)} = \frac{1}{N} \sum_i^N (P_2^{(i)} - P_1^{(i)})$$ [@Wasserman2004].
To provide a concrete example, if the number of trials is 20, $n=m=20$, $X = 11$, and $Y = 4$ Sampling 10000 times from the in the previously described manner yields the histogram presented in Figure \@ref(fig:mon) 
```
```{r mon, fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histogram displaying displaying posterior of delta from Monte Carlo integration"}
p1 <- rbeta(10000, 12, 10)
p2 <- rbeta(10000, 5, 17)
delta <- p2 - p1
delta_df <- data.frame(delta = delta)
hist <- ggplot(delta_df, aes(x=delta)) + 
  geom_histogram(colour="black", fill="#856cb0") +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank())
  #theme(text = element_text(size=12, 
  #                          family = "CMU Serif",
   #                         face = "plain",
   #                         colour = "black",
    #                        hjust = 0.5,
     #                       vjust = 0.5,
      #                      angle = 0,
       #                     lineheight = 0.9))

hist
```

@@

Created at Columbia University by Bob Carpenter and Andrew Gelman, Stan is a domain specific programming language that provides tools for specifying, executing and evaluating statistical models.
Currently comprised of more than 45 core developers from a wide array of educational institutions, private entities, and research organizations, over the past 9 years the Stan project has grown to become one of the leading options for performing Bayesian inference computationally [@Stan2021].
Among the many capabilities of the Stan language, is the ability to define priors, specify dependencies between variables, and impose known limits on parameters.
After all model components have been specified, a Stan model compiles first to C++ and then to binary, which can then be executed to produce an estimation of the posterior densities of all parameters, including intermediate ones, obtained through sampling, with accompanying metadata for evaluating the sampling process.

@@

A typical Stan program is comprised of several, optional *program blocks*, which provide a general organizational structure to a Stan program and generally allow for the declaration of variables and accompanying statements about those variables.
Block types include functions, data, transformed data, parameters, transformed prarameters, model, and generated quantities. 
Each of these block types serve a different purpose in a general modelling schema; however no block is required for the Stan compiler to execute, and an empty string is considered a valid Stan program (though it may not yield a very useful model).
After the program samples the user specified number of iterations and completes execution, a fit object and accompanying summary statistics are produced, which allow for performing post-hoc evaluation such as the posterior predictive check. 
```{r include=FALSE, cache=TRUE}
sharks = read.csv('./data/florida-shark.csv')
```
```{css, echo=FALSE}
.small {
  font-size: small;
}
```

@@

sharks_data = list(
    'N' = length(sharks$population),
    'population'= sharks$population,
    'year'= sharks$year,
    'attacks' = sharks$attacks
)
fit = stan(model_code = sharks_stancode, cores=4, data=sharks_data, chains=4, iter=2000)
```
```{r, cache=TRUE, class.output="small", comment='', tidy=TRUE, size='footnotesize', tidy.opts = list(blank = FALSE, width.cutoff = 30), echo=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)
print(summary(fit, pars=c('alpha', 'beta'))$summary, digits=2)
```
Bob notes that beta, the rate parameter he is interested in understanding, seems to be increasing by nearly 3% each year, although the 97.5% confidence interval of 3.5% causes Bob to wonder how certain he should be in his conclusion about beta. To get a better understanding of the relative frequencies of different samples of the model's parameter values, Bob plots the sample values for each iteration on a histogram, as presented in Figure \@ref(fig:hist).
```{r hist, cache=TRUE, comment='', fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histograms displaying frequencies of varying sampled values for parameters alpha and beta"}
stan_hist(fit, pars=c('alpha', 'beta'), fill = "#856cb0") +#, fill="grey") +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank()) +
  labs(xaxis = c('alpha', 'beta'))
```
The final analysis step for Bob's Bayesian model is ensuring the chains mixed properly, which Bob determines qualitatively using a trace plot as shown in Figure \@ref(fig:trace). 
```{r trace, cache=TRUE, comment='', fig.height=2, echo=FALSE, warning=FALSE, fig.align='center', fig.cap="Each chain represents a sequence of states sampled from the posterior for parameters alpha and beta. Four chains were sampled for this model, and this trace plot is a qualitative method for determining if the sample space for each parameter was sufficiently explored. Here, the sampled value of each parameter (y-axis) is plotted for every iteration Stan evaluated. In this case, Bob specified 2000 total iterations and 1000 burn in iterations, so the first 1000 samples were discarded and not included in the posterior or the trace."}
stan_trace(fit, pars=c('alpha', 'beta')) +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank())
```
As it is clear the chains mixed well, Bob is now relatively confident in the results his model is producing. While he does trust the samples evaluated by his model and his analysis of Figure \@ref(fig:hist)---that the rate of shark attacks is indeed increasing in Florida by some amount---the general lack of concentrated density in both histograms leads Bob to determine that he should collect more data and continue sampling his fit in an attempt to gain a better understanding of preciesly how much shark attacks are increasing each year.
Stan, and more generally Bayesian inference, is a good option for a task like this, because Bob can simply collect more data and continue training this model with his new data, without the need to start the fitting process over again.
